\newglossaryentry{X}
{
  name={\(X\)},
  description={Observed random variable.},
}
\newglossaryentry{bX}
{
  name={\(\bX\)},
  description={Bold symbols indicate sets of variables \((X_{1},\dots,X_{N})\).},
}
\newglossaryentry{x}
{
  name={\(x\)},
  description={Specific value of the random variable \(X\).},
}
\newglossaryentry{bx}
{
  name={\(\bx\)},
  description={Set of observations of the set of variables \(\bX\).},
}
\newglossaryentry{Z}
{
  name={\ensuremath{Z}},
  description={Latent random variable.},
}
\newglossaryentry{P}
{
  name={\ensuremath{P}},
  description={Probability distribution.},
}
\newglossaryentry{Q}
{
  name={\ensuremath{Q}},
  description={Variational probability distribution used in variational inference methods.},
}
\newglossaryentry{P(x)}
{
  name={\ensuremath{P(x)}},
  description={Either probability mass function or density function of \(X\) evaluated on \(x\).},
}
\newglossaryentry{pa(X)}
{
  name={\ensuremath{pa(X)}},
  description={The set of parents of node \(X\) in a graph. The notation \(pa_{X}\) is also used.},
}
\newglossaryentry{pa(x)}
{
  name={\ensuremath{pa(x)}},
  description={Specific value of the random variables conforming the set of parents of node \(X\) in a graph. The notation \(pa_{x}\) is also used. Using lowercase in the random variable is also used in other graph nodes subsets, such as children and co-parents.},
}
\newglossaryentry{cp(X,Y)}
{
  name={\ensuremath{cp(X,Y)}},
  description={Co-parents of \(Y\) with \(X\), that is, \(pa(Y)\backslash \{X\}\).},
}
\newglossaryentry{ne(X)}
{
  name={\ensuremath{ne(X)}},
  description={Neighbor nodes of \(X\) in a graph.},
}
\newglossaryentry{ch(X)}
{
  name={\ensuremath{ch(X)}},
  description={Children nodes of \(X\) in a graph. },
}
\newglossaryentry{theta}
{
  name={\ensuremath{\theta}},
  description={Single parameter of a model \(P(\bx \mid \theta)\).},
}
\newglossaryentry{btheta}
{
  name={\ensuremath{\btheta}},
  description={Set of parameters of a parametric model \(P(\bx \mid \btheta)\).},
}
\newglossaryentry{trace}
{
  name={\ensuremath{tr()}},
  description={Trace of the given matrix.},
}
