
\begin{definition}
A \emph{graph} \(G = (V,E)\) is a set of vertices or nodes \(V\) and edges \(E\subset
V\times V\) between them.
If \(V\) is a set of ordered pairs then the graph is called a \emph{directed
  graph}, otherwise if \(V\) is a set of unordered pairs it is called an \emph{undirected graph}.
\end{definition}

\begin{figure*}[h]
\centering
\begin{tikzpicture}[
  node distance=1cm and 0.5cm,
  mynode/.style={draw,circle,text width=0.5cm,align=center}
]

\node[mynode] (a) {A};
\node[mynode,below right=of a] (b) {B};
\node[mynode,above right=of b] (c) {C};

\node[mynode, right=of c] (d) {A};
\node[mynode,below right=of d] (e) {B};
\node[mynode,above right=of e] (f) {C};

\path (c) edge[-latex] (a)
(a) edge[-latex] (b)
(b) edge[latex-] (c);

\draw (d) -- (e) -- (f) -- (d);

\end{tikzpicture}
\caption{Example of directed and undirected graph, respectively.}
\label{fig:graphs}
\end{figure*}

\begin{definition}
In a directed graph \(G = (V, E)\), a \emph{directed path} \(A \to B\) is a sequence of vertices \({A = A_0,
  A_1,\dots,A_{n-1}, A_n = B}\) where \((A_i, A_{i+1}) \in E \ \forall i \in
0,\dots ,n-1\).

If \(G\) is a undirected graph, \(A \to B\) is an \emph{undirected path} if \(\{A_i, A_{i+1}\} \in E \ \forall i \in
0,\dots, n-1\)
\end{definition}

\begin{definition}
Let \(A,B\) be two vertices of a directed graph \(G\). If \(A \to B\) is a
directed path and \(B \not \to A\) (meaning there isn't a directed path from
\(B\) to \(A\)), then \(A\) is called an \emph{ancestor} of \(B\) and \(B\) is called a \emph{descendant} of \(A\).
\end{definition}

For example, in the figure \ref{fig:graphs}, \(C\) is an ancestor of \(B\).

\begin{definition}
A \emph{directed acyclic graph (DAG)} is a directed graph such that no directed path between any two nodes revisits a vertex.
\end{definition}


\begin{figure}[h]
\centering
\begin{tikzpicture}[
  node distance=1cm and 0.5cm,
  mynode/.style={draw,circle,text width=0.5cm,align=center}
]

\node[mynode] (a) {A};
\node[mynode,below right=of a] (b) {B};
\node[mynode,above right=of b] (c) {C};

\path (c) edge[-latex] (a)
(a) edge[-latex] (b)
(b) edge[-latex] (c);

\end{tikzpicture}
\captionof{figure}{Example of graph which isn't a DAG.}
\label{fig:not_dag}
\end{figure}

As we can see in the figure \ref{fig:not_dag}, \(A \to B \to C \to A \to B\) is a
path from \(A\) to \(B\) that revisits \(A\).

Now where are going to define some relations between nodes in a DAG.

\begin{definition}
The \emph{parents} of a node \(A\) is the set of nodes \(B\) such that there is a
directed edge from \(B\) to \(A\). The same applies for the \emph{children} of a node.

The \emph{Markov blanket} of a node is composed by the node itself, its children, its parents and the parents
of its children.
\end{definition}


\begin{figure}[h]
\centering
\begin{tikzpicture}[
  node distance=1cm and 0.5cm,
  mynode/.style={draw,circle,text width=0.5cm,align=center}
]

\node[mynode] (a) {A};
\node[mynode,below right=of a] (b) {B};
\node[mynode,above right=of b] (c) {C};
\node[mynode,below right=of b] (d) {D};
\node[mynode,below left=of b] (e) {E};
\node[mynode,above right=of d] (f) {F};
\node[mynode, above right=of f] (h) {H};

\path (c) edge[-latex] (a)
(a) edge[-latex] (b)
(b) edge[latex-] (c)
(b) edge[-latex] (e)
(c) edge[-latex] (f)
(b) edge[-latex] (d)
(f) edge[-latex] (d)
(h) edge[-latex] (f)
;

\end{tikzpicture}
\captionof{figure}{Directed acyclic graph}
\label{fig:relations}
\end{figure}

\begin{definition}
In a graph, the \emph{neighbors} of a node are those directly connected
to it.
\end{definition}

We can use figure \ref{fig:relations} to reflect on these definitions. The parents
of \(B\) are \(pa(B) = \{A,C\}\) and its children are \(ch(B) = \{E,D\}\). Taking this into account, its neighbors
are \(ne(B) = \{A,C,E,D\}\) and its Markov blanket is \(\{A,B,C,D,E,F\}\).

\begin{definition}
Let \(G\) be an undirected graph, a \emph{clique} is a maximally connected
subset of vertices. That is, all the members of the clique are connected to each
others and there is no bigger clique that constains another.

Formally, \(S \subset V\) is a \emph{clique} if and only if \(\forall A,B \in S,
\ \{A,B\} \in E\) and \(\nexists C \in V\backslash S\) such that \(\forall A \in
S, \ \{A, C\} \in E \).
\end{definition}
